From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-12 mechanics/paper.tex
Work on Sec 4

@@

\paragraph{The Type of Smooth Functions} Given a pair of cartesian
spaces $X$ and $Y$ we provide a way to construct the type of smooth
invariant smooth functions from $X$ to $Y$:
\begin{displaymath}
  \inferrule*
  {\Theta \vdash X : \mathsf{CartSp} \\
    \Theta \vdash Y : \mathsf{CartSp}}
  {\Theta \vdash C^\infty(X, Y) : *}
\end{displaymath}
with the following interpretation. Since $C^\infty(X,Y)$ is a proper
type, the relational interpetation in this case is a genuine
relation. We relate functions that are invariant under the
diffeomorphism interpretations of $X$ and $Y$ in the current
relational context $\rho$:
\begin{displaymath}
  \begin{array}{@{\hspace{0.05em}}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \sem{C^\infty(X,Y)}^f\theta & = & \{ f : \mathbb{R}^{\sem{X}^f\theta} \to \mathbb{R}^{\sem{Y}^f\theta} \sepbar f\textrm{ smooth} \} \\
    \sem{C^\infty(X,Y)}^f\theta\theta'\rho & = & \{
    \begin{array}[t]{@{}l}
      (f,f') \sepbar \forall \vec{x} \in \mathbb{R}^{\sem{X}^f\theta}.\\
      \quad (\sem{Y}^r\theta\theta'\rho)(f~\vec{x}) = f'~(\sem{X}^r\theta\theta'\rho \vec{x}) \}
    \end{array}
  \end{array}
\end{displaymath}

@@

FIXME: Administrative things: composition, products

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{const} &:& \typeOfCartSp{Y} \to C^\infty(X, Y) \\
    \mathrm{pair} &:& C^\infty(X,Y) \to C^\infty(X,Z) \to C^\infty(X, Y \times Z) \\
    \mathrm{proj_1}&:& C^\infty(X \times Y, X) \\
    \mathrm{proj_2}&:& C^\infty(X \times Y, Y) \\
    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \\
    \mathrm{id} &:& C^\infty(X,X)
  \end{array}
\end{displaymath}

FIXME: vector spacey things

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \underline{\vec{x}} &:& \typeOfCartSp{\mathbb{R}^n\langle 1, 0 \rangle} \\
    0 &:& \forall g \mathord: \mathsf{GL}(n).~\typeOfCartSp{\mathbb{R}^n\langle g, 0 \rangle} \\
    (+) &:&
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 + t_2 \rangle)\\
    \end{array} \\
    (-) &:&
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 - t_2 \rangle)\\
    \end{array} \\
    (*) &:& FIXME \\
    (\cdot) &:& FIXME \\
  \end{array}
\end{displaymath}

FIXME: trigonometic functions, and exponentials.

\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
  \end{array}
\end{displaymath}

FIXME: division, excusing the partiality

\begin{displaymath}
  (/) : \forall 
\end{displaymath}

FIXME: sum and cross

@@

  \medskip
  \textbf{Vector space operations, and dot product}
  \begin{mathpar}
    \inferrule*
    {\vec{x} \in \mathbb{R}^n}
    {\Theta \sepbar \Gamma; \Delta \vdash \vec{x} : \mathbb{R}^n\langle 1, 0 \rangle}

    \inferrule*
    {\Theta \vdash g : \mathsf{GL}(n)}
    {\Theta \sepbar \Gamma; \Delta \vdash 0 : \mathbb{R}^n\langle g, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle g, t_1 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g, t_2 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 + e_2 : \mathbb{R}^n\langle g, t_1 + t_2 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle g, t_1 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g, t_2 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 - e_2 : \mathbb{R}^n\langle g, t_1 - t_2 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle g_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle g_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(g_1)g_2, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(s)\mathit{ortho}_n(o), 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(s)^n, 0 \rangle}
  \end{mathpar}

  \medskip
  \textbf{Transcendental functions}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, t \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~t, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 2\pi z, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sin}~e : \mathbb{R}\langle 1, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 2\pi z, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cos}~e : \mathbb{R}\langle 1, 0 \rangle}
  \end{mathpar}
  FIXME: division, n-ary products (sum, zip, reduce)
  \caption{A surface syntax for smooth functions}
  \label{fig:smooth-terms}
\end{figure*}

\begin{figure}
  \centering
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \elab{z}_\Delta & = & \mathrm{const}~z \\
      \elab{x}_\Delta & = & \mathrm{project}(x, \Delta) \\
    \end{array}
  \end{displaymath}
  \caption{Desugaring of the surface syntax}
  \label{fig:desugar}
\end{figure}

>>>>

Using the type-level natural numbers and their recursion operator, we
can define a type of $n$-ary products of a cartesian space:
\begin{displaymath}
  \mathit{vec} = \lambda n\mathord:\mathsf{Nat}, x\mathord:\mathsf{CartSp}.~\mathit{natrec}~n~\mathbb{R}^0~(\lambda x.~X \times x)
\end{displaymath}
$n$-ary products of cartesian spaces will be useful in
Section~\ref{sec:examples} when we define Lagrangians describing
systems that are generic in the number of particles involved.

@@

\paragraph{The Type of Smooth Functions} Given a pair of cartesian
spaces $X$ and $Y$ we provide a way to construct the type of smooth
invariant smooth functions from $X$ to $Y$:
\begin{displaymath}
  \inferrule*
  {\Theta \vdash X : \mathsf{CartSp} \\
    \Theta \vdash Y : \mathsf{CartSp}}
  {\Theta \vdash C^\infty(X, Y) : *}
\end{displaymath}
with the following interpretation. Since $C^\infty(X,Y)$ is a proper
type, the relational interpretation in this case is a genuine
relation. We relate functions that are invariant under the
diffeomorphism interpretations of $X$ and $Y$ in the current
relational context $\rho$:
\begin{displaymath}
  \begin{array}{@{\hspace{0.05em}}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \sem{C^\infty(X,Y)}^f\theta & = & \{ f : \mathbb{R}^{\sem{X}^f\theta} \to \mathbb{R}^{\sem{Y}^f\theta} \sepbar f\textrm{ smooth} \} \\
    \sem{C^\infty(X,Y)}^f\theta\theta'\rho & = & \{
    \begin{array}[t]{@{}l}
      (f,f') \sepbar \forall \vec{x} \in \mathbb{R}^{\sem{X}^f\theta}.\\
      \quad (\sem{Y}^r\theta\theta'\rho)(f~\vec{x}) = f'~(\sem{X}^r\theta\theta'\rho \vec{x}) \}
    \end{array}
  \end{array}
\end{displaymath}

@@

Our first set of combinators constructs constant and identity smooth
functions, and composes smooth functions between cartesian spaces. We
also have combinators for pairing and projection for the product of
cartesian spaces.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{const} &:& \typeOfCartSp{Y} \to C^\infty(X, Y) \\
    \mathrm{id} &:& C^\infty(X,X) \\
    (\comp) &:& C^\infty(X,Y) \to C^\infty(Y,Z) \\
    \mathrm{pair} &:& C^\infty(X,Y) \to C^\infty(X,Z) \to C^\infty(X, Y \times Z) \\
    \mathrm{proj_1}&:& C^\infty(X \times Y, X) \\
    \mathrm{proj_2}&:& C^\infty(X \times Y, Y) \\
  \end{array}
\end{displaymath}

Next, we include combinators for constant, invariant, vectors, the
zero vector, vector addition, subtraction and scaling. We also include
the Euclidean dot product of vectors. The type dot product operation
is where the orthogonal group kind $\mathsf{O}(n)$ is used: the dot
product of vectors is invariant under orthogonal transformations.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \underline{\vec{c}} &:& \typeOfCartSp{\mathbb{R}^n\langle 1, 0 \rangle} \\
    0 &:& \forall g \mathord: \mathsf{GL}(n).~\typeOfCartSp{\mathbb{R}^n\langle g, 0 \rangle} \\
    (+) &:&
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 + t_2 \rangle)\\
    \end{array} \\
    (-) &:&
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(n), t_1,t_2 \mathord: \mathsf{T}(n). \\
      \hspace{1.5cm} C^\infty(\mathbb{R}^n\langle g, t_1 \rangle \times \mathbb{R}^n\langle g, t_2 \rangle, \mathbb{R}^n\langle g, t_1 - t_2 \rangle)\\
    \end{array} \\
    (*) &:&
    \begin{array}[t]{@{}l}
      \forall g_1 \mathord: \mathsf{GL}(1), g_2 \mathord: \mathsf{GL}(n).\\
      \quad C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}^n\langle g_2, 0\rangle, \mathbb{R}\langle \mathit{scale}_n(g_1)g_2, 0 \rangle)
    \end{array}
    \\
    (\cdot) &:& 
    \begin{array}[t]{@{}l}
      \forall g \mathord: \mathsf{GL}(1), o \mathord: \mathsf{O}(n).~\\
      C^\infty(\mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0 \rangle \times \\
      \quad\quad\; \mathbb{R}^n\langle (\mathit{scale}_n~g)(\mathit{ortho}_n~o), 0\rangle, \mathbb{R}\langle (\mathit{scale}_n~g)^n, 0 \rangle)
    \end{array}
  \end{array}
\end{displaymath}
Each of these combinators has a straightforward interpretation as a
smooth function, and the proof that the fundamental theorem of logical
relations is maintained for our system (i.e., that
Property~\ref{property:semantic-terms} is maintained), follows
directly from elementary facts about linear algebra.

It will be useful for our examples to have the following trigonometric
functions, exponential function and division operation. The types of
each of these functions describes some of their invariance properties,
for example the periodicity of the $\sin$ and $\cos$ functions. A
technical problem with our current system is that we do not account
for the non-definedness of division at $0$. However, division will be
useful in many of our examples, so we take a pragmatic approach, in
common with most physics textbooks, and informally treat division as
total function.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \sin &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \cos &:& \forall z \mathord: \mathsf{Z}.~C^\infty(\mathbb{R}\langle 1, \underline{2\pi} * z \rangle, \mathbb{R} \langle 1, 0 \rangle) \\
    \mathrm{exp} &:& \forall t \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, t \rangle, \mathbb{R}\langle \mathit{exp}~t, 0 \rangle) \\
    (/) &:& \forall g_1, g_2 \mathord: \mathsf{GL}(1).~C^\infty(\mathbb{R}\langle g_1, 0 \rangle \times \mathbb{R}\langle g_2, 0\rangle, \mathbb{R}\langle g_1g_2^{-1}, 0 \rangle)
  \end{array}
\end{displaymath}

Finally, we include two primitive combinators for dealing with $n$-ary
products of vectors. Both of these combinators will be useful when
defining Lagrangians that are generic in the number of particles.
\begin{displaymath}
  \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathrm{sum} &:& \forall n \mathord: \mathsf{Nat}, g \mathord: \mathsf{GL}(1). C^\infty(\mathit{vec}~n~(\mathbb{R}\langle g, 0 \rangle), \mathbb{R}\langle g, 0\rangle) \\
    \mathrm{cross} &:& 
    \begin{array}[t]{@{}l}
      \forall m,n \mathord: \mathsf{Nat}.\\
      \quad C^\infty(X \times Y, Z) \to \\
      \quad\quad C^\infty(\mathit{vec}~m~X \times \mathit{vec}~n~Y, \mathit{vec}~(m*n)~Z)
    \end{array}
  \end{array}
\end{displaymath}
The combinator $\mathrm{sum}$ sums the list of real numbers it is
given, while $\mathrm{cross}$ takes the cartesian product of two
vectors of elements of cartesian spaces, and maps the given smooth
function over them. The multiplication of natural numbers $m*n$ is
straightforwardly defined in terms of the $\mathit{natrec}$ recursion
operator. We will use $\mathrm{cross}$ in Section~\ref{sec:examples}
to define Lagrangians in terms of the interactions between systems of
particles.

@@

  \medskip
  \textbf{Vector space operations, and dot product}
  \begin{mathpar}
    \inferrule*
    {\vec{c} \in \mathbb{R}^n}
    {\Theta \sepbar \Gamma; \Delta \vdash \vec{c} : \mathbb{R}^n\langle 1, 0 \rangle}

    \inferrule*
    {\Theta \vdash G : \mathsf{GL}(n)}
    {\Theta \sepbar \Gamma; \Delta \vdash 0 : \mathbb{R}^n\langle G, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle G, T_1 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G, T_2 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 + e_2 : \mathbb{R}^n\langle G, T_1 + T_2 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle G, T_1 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G, T_2 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 - e_2 : \mathbb{R}^n\langle G, T_1 - T_2 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}\langle G_1, 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle G_2, 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 * e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G_1)G_2, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 : \mathbb{R}^n\langle \mathit{scale}_n(G)\mathit{ortho}_n(O), 0 \rangle \\
      \Theta \sepbar \Gamma; \Delta \vdash e_2 : \mathbb{R}^n\langle \mathit{scale}_n(G)\mathit{ortho}_n(O), 0 \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash e_1 \cdot e_2 : \mathbb{R}\langle\mathit{scale}_n(G)^n, 0 \rangle}
  \end{mathpar}

  \medskip
  \textbf{Transcendental functions}
  \begin{mathpar}
    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, T \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{exp}~e : \mathbb{R}\langle \mathit{exp}~T, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, \underline{2\pi} * Z \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{sin}~e : \mathbb{R}\langle 1, 0 \rangle}

    \inferrule*
    {\Theta \sepbar \Gamma; \Delta \vdash e : \mathbb{R}\langle 1, \underline{2\pi} * Z \rangle}
    {\Theta \sepbar \Gamma; \Delta \vdash \mathrm{cos}~e : \mathbb{R}\langle 1, 0 \rangle}
  \end{mathpar}
  FIXME: division, n-ary products (sum, zip, reduce)
  \caption{A surface syntax for smooth functions}
  \label{fig:smooth-terms}
\end{figure*}

\begin{figure}
  \centering
  \begin{displaymath}
    \begin{array}{l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
      \elab{z}_\Delta & = & \mathrm{const}~z \\
      \elab{x}_\Delta & = & \underline{\mathrm{project}_\Delta}(x) \\
      \elab{\mathrm{let}~\vec{x} = e_1\textrm{ in }e_2}_\Delta & = & \mathrm{pair}~\mathrm{id}~\elab{e_1}_\Delta \comp \elab{e_2}_{\Delta, \vec{x}} \\
      \elab{(e_1,...,e_n)}_\Delta &=& \underline{\mathrm{pair}_n}(\elab{e_1}_\Delta, ..., \elab{e_n}_\Delta) \\
      \elab{e_1(e_2)}_\Delta &=& \elab{e_2}_\Delta \comp e_1 \\
      \elab{\vec{c}}_\Delta &=& 
    \end{array}
  \end{displaymath}
  \caption{Desugaring of the surface syntax}
  \label{fig:desugar}
\end{figure}
